---
title: "Moneyball"
author: "Jorge Ferreira"
output:
  html_document: default
---

To tackle this challenge, we are provided with two datasets: players historical stats and salaries. Let's take a look at them.

```{r warning = FALSE, message = FALSE}
library(readr)
batting <- read.csv('~/Desktop/Moneyball/Batting.csv')
head(batting)
salaries <- read.csv('~/Desktop/Moneyball/Salaries.csv')
head(salaries)
```

There are plenty of metrics that help us analysing players' performance but we will be focusing on a few of them only, whicih we still need to calculate:

* [Batting Average](https://en.wikipedia.org/wiki/Batting_average)
* [On Base Percentage](https://en.wikipedia.org/wiki/On-base_percentage)
* [Slugging Percentage](https://en.wikipedia.org/wiki/Slugging_percentage)

```{r warning = FALSE, message = FALSE}
# Batting Average
batting$BA <- batting$H / batting$AB

# On Base Percentage
batting$OBP <- (batting$H + batting$BB + batting$HBP)/(batting$AB + batting$BB + batting$HBP + batting$SF)

# Creating X1B (Singles) for Slugging Percentage calculation
batting$X1B <- batting$H - batting$X2B - batting$X3B - batting$HR

# Slugging Average (SLG)
batting$SLG <- ((1 * batting$X1B) + (2 * batting$X2B) + (3 * batting$X3B) + (4 * batting$HR) ) / batting$AB
```

In 2001, we lost Giambja01, Damonjo01 and Saenzol01, who were at the peak of their performance.
```{r warning = FALSE, message = FALSE, fig.align = "center"}
library(ggplot2)
library(ggExtra)
GDS <- subset(batting,playerID %in% c('giambja01','damonjo01','saenzol01'))
ggplot(data=GDS[GDS$yearID<2001,], aes(x=yearID, y=BA, group=1)) +
  geom_line(stat="identity", color="green4") + xlab("Year") + facet_grid(~playerID) + theme_bw() + removeGrid() + ggtitle("Batting Average (1994-2001)")
ggplot(data=GDS[GDS$yearID<2001,], aes(x=yearID, y=OBP, group=1)) +
  geom_line(stat="identity", color="green4") + xlab("Year") + facet_grid(~playerID) + theme_bw() + removeGrid() + ggtitle("On Base Percentage (1994-2001)")
ggplot(data=GDS[GDS$yearID<2001,], aes(x=yearID, y=SLG, group=1)) +
  geom_line(stat="identity", color="green4") + xlab("Year") + facet_grid(~playerID) + theme_bw() + removeGrid() + ggtitle("Slugging Percentage (1994-2001)")
```

As salary is a key condition in our quest for potential players, is useful to merge both datasets.
```{r warning = FALSE, message = FALSE}
data <- merge(batting,salaries,by=c('playerID','yearID'))
```

Although we have data since `r min(data$yearID)`, we only care about the year of 2001.
```{r warning = FALSE, message = FALSE}
data <- subset(data, data$yearID==2001)
```

How are these players performing? 
```{r warning = FALSE, message = FALSE, fig.align = "center"}
data$lost_players <- ifelse(data$playerID == "giambja01" | data$playerID == "damonjo01" | data$playerID == "saenzol01", "Lost_Players", "Available_Players")
mycolors         <- c("Lost_Players" = "green4", "Available_Players" = "grey87")

ggplot(data = data, aes(x = OBP, y = salary)) +
  geom_point(size = 1, aes(colour = lost_players)) +
  scale_color_manual("Status", values = mycolors) + theme_minimal()
```

In green we can see our lost players and in grey the ones we have available on the market. 
Let's see how the players we lost performed that season
```{r warning = FALSE, message = FALSE}
lost_players <- subset(data,playerID %in% c('giambja01','damonjo01','saenzol01'))
lost_players <- lost_players[,c('playerID','H','X2B','X3B','HR','OBP','SLG','BA','AB')]
head(lost_players)
```

We have a lot of players who have 0% of OBP. It might mean that they haven't had the chance to show their skills as they haven't played yet. The same rational for 100% OBP, meaning they might have played once and were lucky enough to perform too well.

Now we have to find replacement players for the key three players we lost! However, you have three constraints:

* The total combined salary of the three players can not exceed 15 million dollars.
* Their combined number of At Bats (AB) needs to be equal to or greater than the lost players.
* Their mean OBP had to equal or be greater than the mean OBP of the lost players

So, am going to highlight potential players who we can consider:

* Will set our max salary at 8M$
* OBP should be higher than `r mean(lost_players$OBP)`
* Will look at the ones whose AB is around 500 (`r sum(lost_players$AB)`/3)

```{r warning = FALSE, message = FALSE, out.width=c('50%', '50%'), fig.show='hold'}
avail.players <- subset(data, !(playerID %in% c('giambja01','damonjo01','saenzol01')))
#OBP vs Salary
ggplot(avail.players,aes(x=OBP,y=salary)) + geom_point() + geom_rect(data=avail.players, mapping=aes(xmin=0.367, xmax=0.5, ymin=0, ymax=8000000), color="red", alpha=0) + theme_minimal() + ggtitle("Players OBP vs Salary (2001)")
#Aggregate AB should be above 1469, so let's set individual AB at around 500.
ggplot(avail.players,aes(x=AB,y=salary)) + geom_point() + ggtitle("Players AB vs Salary (2001)") + geom_rect(data=avail.players, mapping=aes(xmin=500, xmax=700, ymin=0, ymax=8000000), color="red", alpha=0) + theme_minimal()
```

By filtering these players and sorting them, we can make a decision who to sign in next:
```{r warning = FALSE, message = FALSE}
library(dplyr)
potential_players <- subset(avail.players, salary<8000000 & OBP>0.367 & AB>500)
potential_players_resume <- potential_players[,c('playerID','OBP','AB','salary','BA','SLG')]
head(arrange(potential_players_resume,desc(OBP),desc(AB),desc(BA),desc(SLG),salary),10)
```

I would chose Heltoto01, Berkmla01 and Gonzalu01!